The \dealii Library, Version 9.6

### 2024-09-23 9.6/paper.tex
Add PM ackn
some text explaining the relicensing effort

\newcommand{\specialword}[1]{\texttt{#1}}
\newcommand{\dealii}{{\specialword{deal.II}}\xspace}
\newcommand{\pfrst}{{\specialword{p4est}}\xspace}
\newcommand{\trilinos}{{\specialword{Trilinos}}\xspace}
\newcommand{\aspect}{\specialword{Aspect}\xspace}
\newcommand{\petsc}{\specialword{PETSc}\xspace}
\newcommand{\snes}{{\specialword{SNES}}\xspace}
\newcommand{\ts}{{\specialword{TS}}\xspace}
\newcommand{\petscsf}{{\specialword{SF}}\xspace}
\newcommand{\cmake}{{\specialword{CMake}}\xspace}
\newcommand{\candi}{{\specialword{candi}}\xspace}
\newcommand{\MPI}{{\specialword{MPI}}\xspace}
\newcommand{\MPIx}{{\specialword{MPI+X}}\xspace}
\newcommand{\sundials}{{\specialword{SUNDIALS}}\xspace}
\newcommand{\kinsol}{{\specialword{KINSOL}}\xspace}
\newcommand{\ida}{{\specialword{IDA}}\xspace}
\newcommand{\arkode}{{\specialword{ARKODE}}\xspace}
\newcommand{\boost}{{\specialword{Boost}}\xspace}
\newcommand{\kokkos}{{\specialword{Kokkos}}\xspace}

@@

\color{red}
\begin{itemize}
  \item
    Motivation for license change, goals: make contributions from industry
    possible / improve deal.II adoption. Mention reasons against LGPL 2.1,
    Explaing why GPL-3/LGPL-3 is not a good fit. Explain why Apache 2.0
    with LLVM exception.
  \item
    Successful examples LLVM and Kokkos (with same license)
  \item
    Status of external collaborators
  \item
    Approach: Require all new contributions to be dual licensed
\end{itemize}
\color{black}

@@

\todo[inline]{Peter: Anything to list here?}

>>>>

\newcommand{\specialword}[1]{\texttt{#1}}
\newcommand{\dealii}{{\specialword{deal.II}}\xspace}
\newcommand{\pfrst}{{\specialword{p4est}}\xspace}
\newcommand{\trilinos}{{\specialword{Trilinos}}\xspace}
\newcommand{\aspect}{\specialword{Aspect}\xspace}
\newcommand{\petsc}{\specialword{PETSc}\xspace}
\newcommand{\snes}{{\specialword{SNES}}\xspace}
\newcommand{\ts}{{\specialword{TS}}\xspace}
\newcommand{\petscsf}{{\specialword{SF}}\xspace}
\newcommand{\cmake}{{\specialword{CMake}}\xspace}
\newcommand{\candi}{{\specialword{candi}}\xspace}
\newcommand{\MPI}{{\specialword{MPI}}\xspace}
\newcommand{\MPIx}{{\specialword{MPI+X}}\xspace}
\newcommand{\sundials}{{\specialword{SUNDIALS}}\xspace}
\newcommand{\kinsol}{{\specialword{KINSOL}}\xspace}
\newcommand{\ida}{{\specialword{IDA}}\xspace}
\newcommand{\arkode}{{\specialword{ARKODE}}\xspace}
\newcommand{\boost}{{\specialword{Boost}}\xspace}
\newcommand{\kokkos}{{\specialword{Kokkos}}\xspace}
\newcommand{\llvm}{{\specialword{LLVM}}\xspace}

@@

The \dealii project is in the process of relicensing the library under the
terms of the \emph{Apache License 2.0 with LLVM
Exception} \cite{Apache-2.0,LLVM-Exception}. To this end we require all new
code contributions to be dual licensed under the current license (GNU
Lesser General Public License v2.1 \cite{LGPL-2.1}) and the new license
(Apache-2.0 with LLVM-exception).

We have chosen to pursue a relicensing effort for deal.II because of some
long standing problems with the LGPL v2.1 license. Most notably, despite
our clear intent that the strong copyleft principle should only apply to
deal.II source code and not user projects (which should be free to choose
their own license freely) this is not necessarily the case for LGPL v2.1
due to the language used in the library. This makes potential industry
partners that evaluate the use of deal.II for their own projects nervous.
Secondly, our previous choice to use an open source license with strong
copyleft was motivated by our desire to receive back code contributions
from third parties, but it is our observation over the last 10 years that
it hardly helps achieving this goal: Copyleft only applies when a derived
software is sold or published. But almost all of our code contributions are
voluntary and come from individual contributors with an academic
background, where these considerations do not apply.

We have thus decided to switch away from the LGPL v2.1 license towards a
more permissible license. We settled on the Apache License 2.0 with
LLVM-exception in large parts because it is (a) a permissible open source
license with patent clause, (b) considered to be a verbose, well-written
license, and (c) has been recently chosen by other large open source
projects, notably \llvm{} and \kokkos{}.

As a first step we now require all new code contributions to be dual
licensed under the old and new licenses. As a second step we are now in the
process of contacting more than 200 contributors with copyrighted
contributions to the library. As of September 2024 roughly 80\% of the
commits and 80\% of the 1.88 million source lines of code have been
relicensed.

@@

P.~Munch acknowledges the funding by the Swedish Research Council (VR) under grant 2021-04620 and the strategic funding from the IT department of Uppsala University.
